The Concurrent Calculi Formalisation Benchmark

### 2024-02-02 challenges/linearity.tex
Edits and comments
Further edits

\subsubsection{Syntax.}
Our type system does not type processes directly, but instead focuses on the channels used in the process.
The syntax of \emph{session types} \( S, T \), unrestricted type contexts \( \Gamma \) and linear type contexts \( \Delta \) is as follows:
\[
  \begin{array}{r@{\qquad}c@{\qquad}l}
  S,T & \Coloneqq & \Tend \quad\mid\quad \Tbase \quad\mid\quad \Tin{S} \quad\mid\quad \Tout{S} \\
    \Gamma & \Coloneqq & \Cempty \quad\mid\quad \Gamma, l \\
                    \Delta &::= & \Cempty \quad\mid\quad \Cadd{\Delta}{\hastype{x}{S}}
  \end{array}
\]
The \emph{end type} \( \Tend \) describes an endpoint on which no further interaction is possible.
The \emph{base type} \( \Tbase \) describes base values.
The \emph{input type} \( \Tin{S} \) describes endpoints used for receiving a value and then according to \( S \).
The \emph{output type} \( \Tout{S} \) describes endpoints used for sending a value and then according to \( S \).

Typing contexts gather type information about names and variables.
\emph{Unrestricted} contexts are simply sets of names since we only have one
base type. \emph{Linear contexts} associate a type to endpoints. We use
the comma as split/union, overloaded to singletons, and \( \Cempty \) as the
empty context. We extend the Barendregt convention to contexts, so that all
entries are distinct.  Note that the order in which information is added to a
type context does not matter.

@@

\subsubsection{Challenge.}
The objective of this challenge is to prove subject reduction and type safety for our calculus with session types. We start with some lemmas:

>>>>

\subsubsection{Syntax.}
Our type system does not type processes directly, but instead focuses on the channels used in the process.
The syntax of \emph{session types} \( S, T \), unrestricted typing contexts \( \Gamma \) and linear typing contexts \( \Delta \) is as follows:
\[
  \begin{array}{r@{\qquad}c@{\qquad}l}
  S,T & \Coloneqq & \Tend \quad\mid\quad \Tbase \quad\mid\quad \Tin{S} \quad\mid\quad \Tout{S} \\
    \Gamma & \Coloneqq & \Cempty \quad\mid\quad \Gamma, l \\
                    \Delta &::= & \Cempty \quad\mid\quad \Cadd{\Delta}{\hastype{x}{S}}
  \end{array}
\]
The \emph{end type} \( \Tend \) describes an endpoint over which no further interaction is possible.
The \emph{base type} \( \Tbase \) describes base values.
The \emph{input type} \( \Tin{S} \) describes endpoints used for receiving a value and then according to \( S \).
The \emph{output type} \( \Tout{S} \) describes endpoints used for sending a value and then according to \( S \).

Typing contexts gather type information about names and variables.
\emph{Unrestricted} contexts are simply sets of names since we only have one
base type. \emph{Linear contexts} associate a type to endpoints. We use
the comma as split/union, overloaded to singletons, and \( \Cempty \) as the
empty context. We extend the Barendregt convention \todo{Source?} to contexts, so that all
entries are distinct.  Note that the order in which information is added to a
type context does not matter.

@@

\subsubsection{Challenge.}
The objective of this challenge is to prove subject reduction and type safety for our calculus with session types. We start with some lemmata:

### 2024-02-02 short-paper/short-paper.tex
Add a few comments.
Further edits

In this spirit, we propose a new collection of benchmark problems specifically designed to address typical issues that arise when mechanising message-passing concurrency using process calculi. \todo{Why just process calculi and not concurrency in general?}
While many efforts to mechanise results about process calculi have been completed (as we will discuss further below), our experience is that choosing appropriate proof mechanisation techniques and tools remains a significant challenge with a limited understanding of the trade-offs.
This often leads researchers toward a trial-and-error approach, resulting in sub-optimal and ad-hoc solutions.
In turn, this increases the overall effort required for proof mechanisation, and hampers the reuse of techniques and results.

@@

It is therefore our view that the research community in message-passing concurrency (and more specifically, process calculi) will benefit from a standardised set of challenge problems.
To this end, our benchmark problems (discussed in detail in \cref{app:challenges}) are designed to focus on three key aspects which typically cause difficulties when mechanising concurrency theory: \emph{linearity}, \emph{scope extrusion}, and \emph{coinductive reasoning} for infinite behaviours.
The mechanisation of novel research results often requires addressing more than just one of these aspects at once; we see the combination of techniques as a next step beyond the scope of this challenge, as discussed further in \Cref{sec:going-beyond}.
These are certainly not the only factors that may cause issues, but they are fundamental to concurrency theory, and emerge in most mechanisations, as we will see presently while discussing the issues presented by the three key aspects.

\paragraph{Linearity.}
In mechanised meta-theory, addressing linearity means choosing an appropriate representation of the linear context.
The crux of the matter is \emph{context splitting}, as showcased by the typing rule for parallel composition.
While a linear context is perhaps best seen as a multi-set, most proof assistants have better support for list processing.
The latter representation is intuitive, but may require establishing a large number of technical lemmata that are orthogonal to the mathematics of the problem under study, say proving type preservation for session types.
In this space, several designs are possible: one can label occurrences of resources to constrain their usage (\eg~\cite{CicconeP20}) or \alb{something --- ask Francisco} as in \cite{Castro2020}.
Some authors instead impose a multiset structure over lists (\eg~\cite{ChaudhuriLR19,Danielsson12}); still, the effort required to develop the infrastructure is significant.

@@

% - Design of the benchmark
%   - We first list some general design remarks, then for each challenge
%   - General remarks
%     - We do not aim to be comprehensive (list other topics)
%     - However, we believe the best practices developed could also be of use in those topics
%     - We base our challenges on basic textbook theory with interesting proof techniques
%     - The challenges are independent (maybe note ``failure'' of POPLMark that everyone only did the first part)
%     - The challenges are as minimal as possible
%   - Linearity
%     - Type preservation for session types
%     - The main issue is context splitting
%     - Session types are pervasive in the field and simple to understand
%     - Binding two names makes duality obvious
%     - Brief overview of approaches and issues with them
%   - Scope extrusion
%     - Process calculi have several binders, but only restriction is different from ``normal'' ones
%     - The main issue is scope extrusion
%     - The pros and cons of established techniques for handling binders are different for process calculi, and it is unclear how exactly
%     - Brief overview of approaches and issues with them
%   - Coinduction
%     - Strong barbed bisimulation and congruence
%     - The main issue is the encoding of coinductive up-to techniques
%     - Strong equivalences to simplify (we don't have to abstract over internal transitions)
%     - Why barbed equivalences?
%     - No delegation to simplify
%     - Brief overview of approaches and issues with them
%   - Evaluation criteria
%     - We care not only about solutions, but about experiences and comparisons
%     - Mechanisation overhead
%     - Adequacy of the formal statements
%     - Cost of entry for tools and techniques
%     - Slight changes are OK if they are justified and adequate
\section{Design of the Benchmark}\label{sec:design-discussion}
\todo{Add references to support our claims and decisions in this section.}

@@

\subsubsection{Linearity.}
For our challenge on linear reasoning we have chosen a type preservation theorem
for a session type system, which allows us to introduce linear
reasoning with few definitions. Accordingly, we have chosen not to
include channel delegation and to use a reduction instead of a
transition system semantics.
Linearity is of course interesting in many other sub-fields of programming languages, but we have chosen to base our challenge on a session type system primarily due to the common use of session types in concurrency theory.\alb{This is weak and somewhat tautological, as a motivation}
Inspired by Vasconcelos~\cite{Vasconcelos2012}, we use a syntax where
restriction binds two names. This presentation makes the duality
of the two names obvious in the type system.
We have chosen a simple notion of well-formedness for our challenge,
and do not, for instance, consider deadlocks. While more
sophisticated notions of well-formedness are interesting, the
proofs of these properties would complicate the challenge, distracting from the linear aspects.

\subsubsection{Scope extrusion.}
This is the challenge most closely related to the POPLMark
challenge since it concerns the properties of binders in restrictions.\alb{I would argue also wrt input}
It is not obvious that the techniques used to solve the original challenges about binders can be easily extended to reasoning about scope extrusion.\alb{I would say not obvious what the pros and cons are, since in fact those techniques (DB, locally nameless, HOAS, nominal) have been used.}
For example, techniques like higher-order abstract syntax, or well-scoped
de\ Bruijn indices do not have the concept of free variables and thus encoding this challenge with them requires some attention.

\subsubsection{Coinduction.}
Our coinduction challenge concerns strong barbed bisimulation and congruence.\alb{we motivate why strong vs weak, but not why barbed}
Though weak barbed congruence is a more common behavioural equivalence for \(\pi\)-terms,\alb{? why terms?} here we prefer strong equivalences to avoid the need to abstract over the number of internal transitions, thus simplifying the theory.
We have chosen not to include delegation in the coinduction challenge since it is orthogonal to our primary aim of exploring coinductive proof techniques.

>>>>

In this spirit, we propose a new collection of benchmark problems specifically designed to address typical issues that arise when mechanising formal models of concurrent and distributed programming languages in general, and process calculi in particular. 
While many efforts to mechanise results about such formalisms have been completed (as we will discuss further below), our experience is that choosing appropriate proof mechanisation techniques and tools remains a significant challenge with a limited understanding of their respective trade-offs.
This often leads researchers toward a trial-and-error approach, resulting in sub-optimal and ad-hoc solutions.
In turn, this increases the overall effort required for proof mechanisation, and hampers the reuse of techniques and results.

@@

It is therefore our view that the research community in concurrency will benefit from a standardised set of challenge problems.
To this end, our benchmark problems (discussed in detail in \cref{app:challenges}) are designed to focus on three key aspects which typically cause difficulties when mechanising concurrency theory: \emph{linearity}, \emph{scope extrusion}, and \emph{coinductive reasoning} for infinite behaviours.
The mechanisation of novel research results often requires addressing more than just one of these aspects at once; we see the combination of techniques as a next step beyond the scope of this challenge, as discussed further in \Cref{sec:going-beyond}.
These are certainly not the only factors that may cause issues, but they are fundamental to concurrency theory, and emerge in most mechanisations, as we will see presently while discussing the issues presented by the three key aspects.

\todo{These paragraphs are still quite messy and unfocused.}
\paragraph{Linearity.}
In mechanised meta-theory, addressing linearity means choosing an appropriate representation of the linear context.
The crux of the matter is \emph{context splitting}, as showcased by the typing rule for parallel composition.
While a linear context is perhaps best seen as a multi-set, most proof assistants have better support for list processing.
The latter representation is intuitive, but may require establishing a large number of technical lemmata that are orthogonal to the mathematics of the problem under study, say proving type preservation for session types.
In this space, several designs are possible: one can label occurrences of resources to constrain their usage (\eg~\cite{CicconeP20}) or \alb{something --- ask Francisco} as in \cite{Castro2020}.
Some authors instead impose a multiset structure over lists (\eg~\cite{ChaudhuriLR19,Danielsson12}); still, the effort required to develop the infrastructure is significant.

@@

% - Design of the benchmark
%   - We first list some general design remarks, then for each challenge
%   - General remarks
%     - We do not aim to be comprehensive (list other topics)
%     - However, we believe the best practices developed could also be of use in those topics
%     - We base our challenges on basic textbook theory with interesting proof techniques
%     - The challenges are independent (maybe note ``failure'' of POPLMark that everyone only did the first part)
%     - The challenges are as minimal as possible
%   - Linearity
%     - Type preservation for session types
%     - The main issue is context splitting
%     - Session types are pervasive in the field and simple to understand
%     - Binding two names makes duality obvious
%     - Brief overview of approaches and issues with them
%   - Scope extrusion
%     - Process calculi have several binders, but only restriction is different from ``normal'' ones
%     - The main issue is scope extrusion
%     - The pros and cons of established techniques for handling binders are different for process calculi, and it is unclear how exactly
%     - Brief overview of approaches and issues with them
%   - Coinduction
%     - Strong barbed bisimulation and congruence
%     - The main issue is the encoding of coinductive up-to techniques
%     - Strong equivalences to simplify (we don't have to abstract over internal transitions)
%     - Why barbed equivalences?
%     - No delegation to simplify
%     - Brief overview of approaches and issues with them
%   - Evaluation criteria
%     - We care not only about solutions, but about experiences and comparisons
%     - Mechanisation overhead
%     - Adequacy of the formal statements
%     - Cost of entry for tools and techniques
%     - Slight changes are OK if they are justified and adequate
\section{Design of the Benchmark}\label{sec:design-discussion}
\todo{Add references to support our claims and decisions in this whole section.}

@@

\subsubsection{Linearity.}
For our challenge on linear reasoning we have chosen a type preservation theorem
for a system using session types, whose use are common in concurrency theory. This allows us to introduce linear
reasoning with few definitions. Accordingly, we omit channel delegation and use a reduction instead of a
transition system semantics.
Inspired by Vasconcelos~\cite{Vasconcelos2012}, we define a syntax where
restriction binds two names, which highlights their duality within the type system.
We have chosen a simple notion of well-formedness,
and do not, for instance, consider deadlocks. While more
sophisticated notions of well-formedness are interesting, the
proofs of these properties would distract from the core linear aspects of the challenge.

\subsubsection{Scope extrusion.}
This is the challenge most closely related to the POPLMark
challenge since it concerns the properties of binders in restrictions and input. 
It is not obvious what the limitations and advantages are in leveraging the techniques used to solve the original challenges to reason about scope extrusion. 
For example, techniques like higher-order abstract syntax or well-scoped
de\ Bruijn indices lack the concept of free variables and thus encoding this challenge with them requires attention.

\subsubsection{Coinduction.}
Our coinduction challenge concerns strong barbed bisimulation and congruence.\alb{we motivate why strong vs weak, but not why barbed}
Though weak barbed congruence is a more common behavioural equivalence for \(\pi\)-terms,\alb{? why terms?} here we prefer strong equivalences to avoid the need to abstract over the number of internal transitions, thus simplifying the theory.
We exclude delegation from the coinduction challenge since it is orthogonal to our primary aim of exploring coinductive proof techniques.
